Normalization by Evaluation for Sized Dependent Types

### 2017-02-25 nbe/icfp17.tex
[ ICFP 17 ] proof* without \qed

\begin{theorem}[Reflection and reification]
\label{thm:rere}
Let $T \in \SET[\ell]$ and $T \shape S_1$ and $T \shape S_2$.
\begin{enumerate}
\item If $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
\item If $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
\end{enumerate}
\end{theorem}
\begin{proof}
By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
\begin{caselist}

@@

\nextcase $T \evalsto \forallT B$
\begin{gather*}
  \ru{B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T \in \SET[\ell]}
\\[2ex]
  \ru{S_1 \evalsto \forallT{B_1} \qquad
      B[\alpha] \shape B_1[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_1}
\\[2ex]
  \ru{S_2 \evalsto \forallT{B_2} \qquad
      B[\alpha] \shape B_2[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $\alpha_1,\alpha_2 \in \SIZE$.
  Since $n_1 \ann{\alpha_1} = n_2 \ann{\alpha_2} \in \NE$ by Lemma~\ref{lem:closne}, we obtain
  $\up{B_1[\alpha_1]}{(n_1 \ann{\alpha_1})} = \up{B_2[\alpha_2]}{(n_2 \ann{\alpha_2})} \in B[\alpha_1]$
  by induction hypothesis.
  Thus, $(\up{S_1}{n_1}) \ann{\alpha_1} = (\up{S_2}{n_2}) \ann{\alpha_2} \in B[\alpha_1]$
  by weak head expansion, which entails the goal by definition of $\EL(T)$.
\item Assume $k \in \NN$ and note that $\lev k = \lev k \in \SIZE$, hence,
  $t_1\,\ann{\lev k} = t_2\,\ann{\lev k} \in B[\lev k]$.  Thus, by induction hypothesis,
  $\R{k+1}{\down{B_i[\lev k]}(t_i\,\ann{\lev k})}{v_i}$, and finally
  $\R k {\down{S_i} t_i} {\lambda v_i}$ by definition of read back.

\end{enumerate}

\end{caselist}
\end{proof}

@@

The eliminations for natural numbers are acurrately modeled.
\begin{lemma}[Case]
  \label{lem:case}
If $a = a' \in \Nat (\alpha+1)$
and $B = B' \in \Nat (\alpha+1) \to \Set[\ell]$
and $f_z = f_z' \in B\,(\zero \beta)$
and $f_s = f_s' \in \funT x {\Nat \alpha} B\,(\suc \gamma x)$
then $a\,\case[\ell] B {f_z}{f_s} = a'\,\case[\ell] {B'} {f_z'}{f_s'} \in B\,a$.
\end{lemma}
\begin{proof}
By induction on $a = a' \in \Nat(\alpha+1)$.
\begin{caselist}

@@

\nextcase $a \evalsto \up T n$ and $a' \evalsto \up {T'} n'$ with $n = n' \in \NE$.
Let $D = \down{\Nat \infty \to \Set[\ell]} B$
and $d_z = \down{B\,(\zero \infty)} f_z$
and $d_s = \down{\funS x {\Nat\,\infty} {B\,(\suc \infty x)}} f_s$
and $e = \case[\ell] D {d_z}{d_s}$.
% $e = \case[\ell]
%   {(\down{\Nat \infty \to \Set[\ell]} B)}
%   {(\down{B\,(\zero \infty)} f_z)}
%   {(\down{\funS x {\Nat\,\infty} {B\,(\suc \infty x)} f_s)}$
Let $D',d_z',d_s',e'$ be defined analogously from $B',f_z',f_s'$.  It suffices to show $e = e' \in \ELIM$, since then we have $n\,e = n'\,e' \in \NE$ by the closure properties of $\NE$ (Lemma~\ref{lem:closne}),
and $\up{B[a]} (n\,e) = \up{B'[a']} (n'\,e') \in B[a]$ by reflection (Theorem~\ref{thm:rere}).
The remaining goal $\case[\ell] D {d_z}{d_s} = \case[\ell] {D'} {d_z'}{d_s'} \in \ELIM$ follows by the closure properties for eliminations (Lemma~\ref{lem:closelim}), since $D = D' \in \NF$ and $d_z = d_z' \in \NF$ and $d_s = d_s' \in \NF$ all hold by reification (Theorem~\ref{thm:rere}).
\end{caselist}
\end{proof}

@@

\nextcase $\alpha = \infty$.
Note that $a = a' \in \Nat \alpha$ for some $\alpha < \infty$ by Lemma~\ref{lem:limit}.
By induction hypothesis, $g = g' \in B\,\alpha\,a$.  Since
$B\,\alpha\,a \leq B\,\infty\,a$ by assumption, the goal
$g = g' \in B\,\infty\,a$ follows by subsumption.
\end{caselist}
\end{proof*}

@@

\begin{theorem}[Fundamental theorem]
  \label{thm:fund}
  \bla
  \begin{enumerate}
  \item
  If $\der \Gamma$ then $\models \Gamma$.
  \item
  If $\Gamma \der J$ then % $\models \Gamma$ and
  $\Gamma \models J$
  \end{enumerate}
\end{theorem}
\begin{proof}
Simultaneously, by induction on the derivation.
% The soundness of subtyping and the conversion rule is due to subsumption (lemmata \ref{lem:subsump} and \ref{lem:usubsump}).

@@

We now argue that this choice is possible,
namely that $a \eta, a'\eta, b\rho \in \SIZE$.
Note that the induction hypothesis gives us
$a \eta  = a  \eta' \in \SIZE$ and
$a' \eta = a' \eta' \in \SIZE$.
By resurrection (Lemma~\ref{lem:resenv}) we have $\rho \in \resurrect\Gamma$,
thus, by induction hypothesis, $b \rho \in \SIZE$.
\end{caselist}
\end{proof}

@@

We write $\var\Gamma$ for $\var{|\Gamma|}$.
We define the identity environment $\rho_\Gamma$ by induction on $\Gamma$.
We set $\rho_{\cempty} = \sempty$ and
\[
\begin{array}{lll}
  \rho_{\cext \Gamma T} & = & (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma) \\
  \rho_{\eext \Gamma \Size} & = & (\rho_\Gamma, \var\Gamma) \\
\end{array}
\]
\begin{lemma}[Identity environment]
  If $\der \Gamma$ then
  $\rho_\Gamma \in \Gamma$.
  % $\rho_\Gamma = \rho_\Gamma \eeq \rho_\Gamma \in \Gamma$
\end{lemma}
\begin{proof}
By induction on $\der \Gamma$.
\begin{caselist}

@@

\nextcase
\[
  \ru{\der \Gamma
    }{\der \eext\Gamma \Size}
\]
By induction hypothesis $\rho_\Gamma \in \Gamma$.  Since $\var\Gamma \in \SIZE$ we conclude
$(\rho_\Gamma, \var\Gamma) \in \eext\Gamma \Size)$.
\end{caselist}
\end{proof}

@@

\begin{proof}
We prove the propositions for $A'$ and $T'$ (instead of $A$ and $T$)
simultaneously by induction on $A' \in s$.

@@

\nextcase $A' \evalsto \piT A B$ and $\resurrect\Gamma \der T' = \piT U T$.
\begin{enumerate}
\item
The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
To demonstrate $\LR \Gamma t {T'} {\up {A'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{B[a]}(n\,\down A a)} {B[a]}$.
By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down A a} {U\xi}$, and
together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down A a) : T(\xi,u)$.
The goal follows by induction hypothesis (\ref{it:into}) for
% $B[a] \in s$.
$\LRTS{\Gamma'}{T(\xi,u)}{B[a]} s$.
% This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
% we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
% such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
% Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
% We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
% Thus, it is sufficient to show that there is some normal form $v$ such that
% $\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
% However, this follows by induction hypothesis (\ref{it:out}) on
% assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
\end{enumerate}
\end{caselist}
\end{proof}

@@

\begin{theorem}[Fundamental theorem of typing]
\label{thm:fundty}
\bla
\begin{enumerate}
\item
If $\Gamma \der t : T$ then $\Gamma \valid t : T$.
\item
If $\Gamma \der \sigma : \Delta$ then
$\Gamma \valid \sigma : \Delta$
\end{enumerate}
\end{theorem}
\begin{proof}
Each by induction on the derivation.

@@

\nextcase $\forall$-elimination.
\[
  \ru{\Gamma \der t : \forall T \qquad % \Gamma \der a : \Size \qquad
      \resurrect \Gamma \der a,b : \Size
    }{\Gamma \der t \ann a : T[b]}
\]
Assume $\LRSub {\Delta} {\sigma} {\tau} \Gamma {\eta} {\rho}$ and show
$\LR \Delta {t \ann a \sigma} {T[b]\tau} {t \ann a \eta} {T[b]\rho}$.
By induction hypothesis we have $\LR \Delta {t \sigma} {(\forall T)\tau} {t \eta} {(\forall T) \rho}$.
It suffices to show
\begin{enumerate}
\item $\resurrect\Delta \der a\sigma : \Size$: follows from $\resurrect\Gamma \der a : \Size$
  via $\Delta \der \sigma : \Gamma$ and $\resurrect\Delta \der \sigma : \resurrect\Gamma$
  and substitution.
%\item $\resurrect\Delta \der b\tau : \Size$.
\item $a\eta \in \SIZE$: follows from $\resurrect\Gamma \der a : \Size$
  by Lemma~\ref{lem:wklrsub}, part (\ref{it:sizesub}).
%\item $b\rho \in \SIZE$.
\item $\LRSize {\resurrect \Delta} {b\tau} {b\rho}$: follows from induction hypothesis on $\resurrect \Gamma \der b : \Size$ via $\LRSub {\resurrect \Delta} \tau \tau {\resurrect\Gamma} \rho \rho$ (from Lemma~\ref{lem:wklrsub}).
\end{enumerate}

\end{caselist}
\end{proof}

\begin{lemma}[Identity environment]
  \label{lem:idenv}
  If $\der \Gamma$ then
  $\LRS \Gamma \sid \Gamma \rhog$.
\end{lemma}
\begin{proof}
  By induction on $\der \Gamma$.
\begin{caselist}

\nextcase
\[
  \ru{\der \Gamma \qquad \resurrect\Gamma \der T
    }{\der \cext \Gamma T}
\]
Let $f = \up{T\rhog} \var\Gamma$.
We have to show $\LRS{\cext \Gamma T} \sid {\cext\Gamma T} {(\rhog, f)}$.
Note that $\sid = (\slift,\ind0)$, thus it remains to show that
$\LRS {\cext \Gamma T} {\slift} \Gamma \rhog$ and
$\LR {\cext \Gamma T} {\ind0} {T\slift} f {T\rhog}$.
The first goal follows by weakening from the induction hypothesis
$\LRS \Gamma \sid \Gamma \rhog$.
%Further, by resurrection we get
%$\LRS {\cext \Gamma T} {\slift} \Gamma \rhog$, and by resurrection
%$\LRS {\cext{\resurrect\Gamma}T}{\slift}{\resurrect\Gamma}\rhog$.
Since $\rhog \in \resurrect\Gamma$, we get $T \rhog \in \SET[\ell]$ for some $\ell$
by the first fundamental theorem (Thm.~\ref{thm:fund}).
Thus, the second goal follows by Cor.~\ref{cor:fresh}.
% Further note that for all $\Gamma' = \cext {\cext \Gamma T} {\Delta}$ and
% $\xi : \Gamma' \leq \cext \Gamma T$ we have that
% $\RNe{\Gamma'} {\var\Gamma} {\ind{|\Gamma'|\monus|\Gamma|}} = {\ind{|\Delta|}}$ and
% $\Gamma' \der \ind 0 \xi = \ind{|\Delta|} : T\xi$.
% This allows us to apply Thm.~\ref{thm:into} to get $\ind0$ into the logical relation as
% $\LR {\cext \Gamma T} {\ind0} T f {T\rhog}$.

\nextcase

\end{caselist}
\end{proof}

\begin{corollary}[Soundness of NbE]
  \label{cor:soundnbe} \bla
  \begin{enumerate}
  \item \label{it:sound}
  If\/ $\Gamma \der t : T$ then $\Gamma \der t = \nbe \Gamma T t : T$.
  \item
  If\/ $\Gamma \der t,t' : T$ and $\nbe \Gamma T t \equiv \nbe \Gamma T {t'}$ then
  $\Gamma \der t = t' : T$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  \bla
  \begin{enumerate}

@@

  \item By transitivity of judgmental equality from (\ref{it:sound}).
     Here we use that the computed normal forms are identical because
     we erase all size annotations $\ann\alpha$ to $\ann\infty$ during readback.
  \end{enumerate}
\end{proof}

@@

\begin{lemma}[Semantic subtyping implies algorithmic subtyping]
\label{lem:lrasub}
\bla
   If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
\end{lemma}
\begin{proof}
By induction on % $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
$\LRTS \Gamma {T_1'} {A_1'} {s_1}$ and $\LRTS \Gamma {T_2'} {A_2'} {s_2}$
and cases on $A_1' \leq A_2' \in \TYPE$
we prove $\Gamma \der T_1' \asub T_2'$.

@@

\nextcase $A_i' \evalsto \piT {A_i} {B_i}$ and  $T_i' \evalsto \piT{U_i}{T_i}$.
From $\LRTS \Gamma {U_i} {A_i} {s_i}$ and $A_2 \leq A_1 \in \TYPE$ we get $\Gamma \der U_2 \asub U_1$
by induction hypothesis on $A_i \in s_i$.
Let $a := {\up{A_2} \var\Gamma}$.  With $\LR {\cext \Gamma {U_2}} {\ind 0} {(U_2\slift)} a {A_2}$ we get
$\LRTS {\cext \Gamma {U_2}} {T_i} {B_i[a]} s$ and $B_1[a] \leq B_2[a] \in \TYPE$.  Thus by induction hypothesis on $B_i[a] \in s_i$ we obtain $\cext \Gamma {U_2} \der T_1 \leq T_2$, together $\Gamma \der T_1' \asub T_2'$.

\end{caselist}
\end{proof}

>>>>

\begin{theorem}[Reflection and reification]
\label{thm:rere}
Let $T \in \SET[\ell]$ and $T \shape S_1$ and $T \shape S_2$.
\begin{enumerate}
\item If $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
\item If $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
\end{enumerate}
\end{theorem}
\begin{proof*}
By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
\begin{caselist}

@@

\nextcase $T \evalsto \forallT B$
\begin{gather*}
  \ru{B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T \in \SET[\ell]}
\\[2ex]
  \ru{S_1 \evalsto \forallT{B_1} \qquad
      B[\alpha] \shape B_1[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_1}
\\[2ex]
  \ru{S_2 \evalsto \forallT{B_2} \qquad
      B[\alpha] \shape B_2[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $\alpha_1,\alpha_2 \in \SIZE$.
  Since $n_1 \ann{\alpha_1} = n_2 \ann{\alpha_2} \in \NE$ by Lemma~\ref{lem:closne}, we obtain
  $\up{B_1[\alpha_1]}{(n_1 \ann{\alpha_1})} = \up{B_2[\alpha_2]}{(n_2 \ann{\alpha_2})} \in B[\alpha_1]$
  by induction hypothesis.
  Thus, $(\up{S_1}{n_1}) \ann{\alpha_1} = (\up{S_2}{n_2}) \ann{\alpha_2} \in B[\alpha_1]$
  by weak head expansion, which entails the goal by definition of $\EL(T)$.
\item Assume $k \in \NN$ and note that $\lev k = \lev k \in \SIZE$, hence,
  $t_1\,\ann{\lev k} = t_2\,\ann{\lev k} \in B[\lev k]$.  Thus, by induction hypothesis,
  $\R{k+1}{\down{B_i[\lev k]}(t_i\,\ann{\lev k})}{v_i}$, and finally
  $\R k {\down{S_i} t_i} {\lambda v_i}$ by definition of read back.
\qed
\end{enumerate}

\end{caselist}
\end{proof*}

@@

The eliminations for natural numbers are acurrately modeled.
\begin{lemma}[Case]
  \label{lem:case}
If $a = a' \in \Nat (\alpha+1)$
and $B = B' \in \Nat (\alpha+1) \to \Set[\ell]$
and $f_z = f_z' \in B\,(\zero \beta)$
and $f_s = f_s' \in \funT x {\Nat \alpha} B\,(\suc \gamma x)$
then $a\,\case[\ell] B {f_z}{f_s} = a'\,\case[\ell] {B'} {f_z'}{f_s'} \in B\,a$.
\end{lemma}
\begin{proof*}
By induction on $a = a' \in \Nat(\alpha+1)$.
\begin{caselist}

@@

\nextcase $a \evalsto \up T n$ and $a' \evalsto \up {T'} n'$ with $n = n' \in \NE$.
Let $D = \down{\Nat \infty \to \Set[\ell]} B$
and $d_z = \down{B\,(\zero \infty)} f_z$
and $d_s = \down{\funS x {\Nat\,\infty} {B\,(\suc \infty x)}} f_s$
and $e = \case[\ell] D {d_z}{d_s}$.
% $e = \case[\ell]
%   {(\down{\Nat \infty \to \Set[\ell]} B)}
%   {(\down{B\,(\zero \infty)} f_z)}
%   {(\down{\funS x {\Nat\,\infty} {B\,(\suc \infty x)} f_s)}$
Let $D',d_z',d_s',e'$ be defined analogously from $B',f_z',f_s'$.  It suffices to show $e = e' \in \ELIM$, since then we have $n\,e = n'\,e' \in \NE$ by the closure properties of $\NE$ (Lemma~\ref{lem:closne}),
and $\up{B[a]} (n\,e) = \up{B'[a']} (n'\,e') \in B[a]$ by reflection (Theorem~\ref{thm:rere}).
The remaining goal $\case[\ell] D {d_z}{d_s} = \case[\ell] {D'} {d_z'}{d_s'} \in \ELIM$ follows by the closure properties for eliminations (Lemma~\ref{lem:closelim}), since $D = D' \in \NF$ and $d_z = d_z' \in \NF$ and $d_s = d_s' \in \NF$ all hold by reification (Theorem~\ref{thm:rere}).
\qed
\end{caselist}
\end{proof*}

@@

\nextcase $\alpha = \infty$.
Note that $a = a' \in \Nat \alpha$ for some $\alpha < \infty$ by Lemma~\ref{lem:limit}.
By induction hypothesis, $g = g' \in B\,\alpha\,a$.  Since
$B\,\alpha\,a \leq B\,\infty\,a$ by assumption, the goal
$g = g' \in B\,\infty\,a$ follows by subsumption.
\qed
\end{caselist}
\end{proof*}

@@

\begin{theorem}[Fundamental theorem]
  \label{thm:fund}
  \bla
  \begin{enumerate}
  \item
  If $\der \Gamma$ then $\models \Gamma$.
  \item
  If $\Gamma \der J$ then % $\models \Gamma$ and
  $\Gamma \models J$
  \end{enumerate}
\end{theorem}
\begin{proof*}
Simultaneously, by induction on the derivation.
% The soundness of subtyping and the conversion rule is due to subsumption (lemmata \ref{lem:subsump} and \ref{lem:usubsump}).

@@

We now argue that this choice is possible,
namely that $a \eta, a'\eta, b\rho \in \SIZE$.
Note that the induction hypothesis gives us
$a \eta  = a  \eta' \in \SIZE$ and
$a' \eta = a' \eta' \in \SIZE$.
By resurrection (Lemma~\ref{lem:resenv}) we have $\rho \in \resurrect\Gamma$,
thus, by induction hypothesis, $b \rho \in \SIZE$.
\qed
\end{caselist}
\end{proof*}

@@

We write $\var\Gamma$ for $\var{|\Gamma|}$.
We define the identity environment $\rho_\Gamma$ by induction on $\Gamma$.
We set $\rho_{\cempty} = \sempty$ and
\[
\begin{array}{lll}
  \rho_{\cext \Gamma T} & = & (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma) \\
  \rho_{\eext \Gamma \Size} & = & (\rho_\Gamma, \var\Gamma) \\
\end{array}
\]
\begin{lemma}[Identity environment]
  If $\der \Gamma$ then
  $\rho_\Gamma \in \Gamma$.
  % $\rho_\Gamma = \rho_\Gamma \eeq \rho_\Gamma \in \Gamma$
\end{lemma}
\begin{proof*}
By induction on $\der \Gamma$.
\begin{caselist}

@@

\nextcase
\[
  \ru{\der \Gamma
    }{\der \eext\Gamma \Size}
\]
By induction hypothesis $\rho_\Gamma \in \Gamma$.  Since $\var\Gamma \in \SIZE$ we conclude
$(\rho_\Gamma, \var\Gamma) \in \eext\Gamma \Size)$.
\qed
\end{caselist}
\end{proof*}

@@

\begin{proof*}
We prove the propositions for $A'$ and $T'$ (instead of $A$ and $T$)
simultaneously by induction on $A' \in s$.

@@

\nextcase $A' \evalsto \piT A B$ and $\resurrect\Gamma \der T' = \piT U T$.
\begin{enumerate}
\item
The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
To demonstrate $\LR \Gamma t {T'} {\up {A'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{B[a]}(n\,\down A a)} {B[a]}$.
By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down A a} {U\xi}$, and
together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down A a) : T(\xi,u)$.
The goal follows by induction hypothesis (\ref{it:into}) for
% $B[a] \in s$.
$\LRTS{\Gamma'}{T(\xi,u)}{B[a]} s$.
% This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
% we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
% such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
% Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
% We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
% Thus, it is sufficient to show that there is some normal form $v$ such that
% $\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
% However, this follows by induction hypothesis (\ref{it:out}) on
% assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
\qed
\end{enumerate}
\end{caselist}
\end{proof*}

@@

\begin{theorem}[Fundamental theorem of typing]
\label{thm:fundty}
\bla
\begin{enumerate}
\item
If $\Gamma \der t : T$ then $\Gamma \valid t : T$.
\item
If $\Gamma \der \sigma : \Delta$ then
$\Gamma \valid \sigma : \Delta$
\end{enumerate}
\end{theorem}
\begin{proof*}
Each by induction on the derivation.

@@

\nextcase $\forall$-elimination.
\[
  \ru{\Gamma \der t : \forall T \qquad % \Gamma \der a : \Size \qquad
      \resurrect \Gamma \der a,b : \Size
    }{\Gamma \der t \ann a : T[b]}
\]
Assume $\LRSub {\Delta} {\sigma} {\tau} \Gamma {\eta} {\rho}$ and show
$\LR \Delta {t \ann a \sigma} {T[b]\tau} {t \ann a \eta} {T[b]\rho}$.
By induction hypothesis we have $\LR \Delta {t \sigma} {(\forall T)\tau} {t \eta} {(\forall T) \rho}$.
It suffices to show
\begin{enumerate}
\item $\resurrect\Delta \der a\sigma : \Size$: follows from $\resurrect\Gamma \der a : \Size$
  via $\Delta \der \sigma : \Gamma$ and $\resurrect\Delta \der \sigma : \resurrect\Gamma$
  and substitution.
%\item $\resurrect\Delta \der b\tau : \Size$.
\item $a\eta \in \SIZE$: follows from $\resurrect\Gamma \der a : \Size$
  by Lemma~\ref{lem:wklrsub}, part (\ref{it:sizesub}).
%\item $b\rho \in \SIZE$.
\item $\LRSize {\resurrect \Delta} {b\tau} {b\rho}$: follows from induction hypothesis on $\resurrect \Gamma \der b : \Size$ via $\LRSub {\resurrect \Delta} \tau \tau {\resurrect\Gamma} \rho \rho$ (from Lemma~\ref{lem:wklrsub}).
\qed
\end{enumerate}
\end{caselist}
\end{proof*}

\begin{lemma}[Identity environment]
  \label{lem:idenv}
  If $\der \Gamma$ then
  $\LRS \Gamma \sid \Gamma \rhog$.
\end{lemma}
\begin{proof*}
  By induction on $\der \Gamma$.
\begin{caselist}

\nextcase
\[
  \ru{\der \Gamma \qquad \resurrect\Gamma \der T
    }{\der \cext \Gamma T}
\]
Let $f = \up{T\rhog} \var\Gamma$.
We have to show $\LRS{\cext \Gamma T} \sid {\cext\Gamma T} {(\rhog, f)}$.
Note that $\sid = (\slift,\ind0)$, thus it remains to show that
$\LRS {\cext \Gamma T} {\slift} \Gamma \rhog$ and
$\LR {\cext \Gamma T} {\ind0} {T\slift} f {T\rhog}$.
The first goal follows by weakening from the induction hypothesis
$\LRS \Gamma \sid \Gamma \rhog$.
%Further, by resurrection we get
%$\LRS {\cext \Gamma T} {\slift} \Gamma \rhog$, and by resurrection
%$\LRS {\cext{\resurrect\Gamma}T}{\slift}{\resurrect\Gamma}\rhog$.
Since $\rhog \in \resurrect\Gamma$, we get $T \rhog \in \SET[\ell]$ for some $\ell$
by the first fundamental theorem (Thm.~\ref{thm:fund}).
Thus, the second goal follows by Cor.~\ref{cor:fresh}.
% Further note that for all $\Gamma' = \cext {\cext \Gamma T} {\Delta}$ and
% $\xi : \Gamma' \leq \cext \Gamma T$ we have that
% $\RNe{\Gamma'} {\var\Gamma} {\ind{|\Gamma'|\monus|\Gamma|}} = {\ind{|\Delta|}}$ and
% $\Gamma' \der \ind 0 \xi = \ind{|\Delta|} : T\xi$.
% This allows us to apply Thm.~\ref{thm:into} to get $\ind0$ into the logical relation as
% $\LR {\cext \Gamma T} {\ind0} T f {T\rhog}$.
\qed
\end{caselist}
\end{proof*}

\begin{corollary}[Soundness of NbE]
  \label{cor:soundnbe} \bla
  \begin{enumerate}
  \item \label{it:sound}
  If\/ $\Gamma \der t : T$ then $\Gamma \der t = \nbe \Gamma T t : T$.
  \item
  If\/ $\Gamma \der t,t' : T$ and $\nbe \Gamma T t \equiv \nbe \Gamma T {t'}$ then
  $\Gamma \der t = t' : T$.
  \end{enumerate}
\end{corollary}
\begin{proof*}
\bla
\begin{enumerate}

@@

\item By transitivity of judgmental equality from (\ref{it:sound}).
   Here we use that the computed normal forms are identical because
   we erase all size annotations $\ann\alpha$ to $\ann\infty$ during readback.
\qed
\end{enumerate}
\end{proof*}

@@

\begin{lemma}[Semantic subtyping implies algorithmic subtyping]
\label{lem:lrasub}
\bla
   If\/ $\RG \Gamma T A$ and $\RG \Gamma {T'} {A'}$ and $A \leq A' \in \TYPE$ then $\Gamma \der T \asub T'$.
\end{lemma}
\begin{proof*}
By induction on % $A_1' \in s_1$ and $A_2' \in s_2$ and cases on the inductive relations
$\LRTS \Gamma {T_1'} {A_1'} {s_1}$ and $\LRTS \Gamma {T_2'} {A_2'} {s_2}$
and cases on $A_1' \leq A_2' \in \TYPE$
we prove $\Gamma \der T_1' \asub T_2'$.

@@

\nextcase $A_i' \evalsto \piT {A_i} {B_i}$ and  $T_i' \evalsto \piT{U_i}{T_i}$.
From $\LRTS \Gamma {U_i} {A_i} {s_i}$ and $A_2 \leq A_1 \in \TYPE$ we get $\Gamma \der U_2 \asub U_1$
by induction hypothesis on $A_i \in s_i$.
Let $a := {\up{A_2} \var\Gamma}$.  With $\LR {\cext \Gamma {U_2}} {\ind 0} {(U_2\slift)} a {A_2}$ we get
$\LRTS {\cext \Gamma {U_2}} {T_i} {B_i[a]} s$ and $B_1[a] \leq B_2[a] \in \TYPE$.  Thus by induction hypothesis on $B_i[a] \in s_i$ we obtain $\cext \Gamma {U_2} \der T_1 \leq T_2$, together $\Gamma \der T_1' \asub T_2'$.
\qed
\end{caselist}
\end{proof*}
